---
title: "R Programming Lab Notebook"
subtitle: Knowledge Engineering Lab
author: Sanmitra Dharmavarapu
githuburl: https://github.com/sanmitraD/Knowledge-Engineering-Lab-Andhra-University
output:
  word_document: default
  html_notebook: default
---

github url: [https://github.com/sanmitraD/Knowledge-Engineering-Lab-coursework](https://github.com/sanmitraD/Knowledge-Engineering-Lab-coursework)

# 1. Exploratory data analysis using R
Load the iris dataset and display the names and type of each column. 
Find statistics such as min, max, range, mean, median, variance and standard deviation for each column of data.

**Iris Data Set**
Iris dataset os the best known dataset to be found in pattern recognition literature. The dataset contains 3 classes of 50 instances each, where each class refers to a type of iris plant.

*Attribute Information*
1. sepal length in cm

2. sepal width in cm
3. petal length in cm
4. petal width in cm
5. class
    + Iris Setosa
    + Iris Versicolour
    + Iris Virginica
    

```{r}

# the Iris dataset is an inbulit dataset avalable in RStudio and we can use it using iris identifier 
print(iris)

```

To display the names and types of each column

```{r}
# to display names of each column we use the names functon

print(names(iris))
```

For diplaying the names along with their types we use the lapply function using class as second argument

```{r}
print(lapply(iris,class))
```

*Finding the mean of individual columns*

```{r}
# we can find mean of a column using mean function

# we can refer to individual columns in a datset using $ symbol as showin below

print(mean(iris$Sepal.Length))
print(mean(iris$Sepal.Width))
print(mean(iris$Petal.Length))
print(mean(iris$Petal.Width))

```

*Finding the median of each column*
```{r}
# we can find median of a column using median function


print(median(iris$Sepal.Length))
print(median(iris$Sepal.Width))
print(median(iris$Petal.Length))
print(median(iris$Petal.Width))
```

*Finding maximum value of each column*
```{r}
# we can find maximum of a column using max function


print(max(iris$Sepal.Length))
print(max(iris$Sepal.Width))
print(max(iris$Petal.Length))
print(max(iris$Petal.Width))

```

*Finding minimum value of each column*
returns a vector containing the minimum and maximum of all the given arguments.
```{r}
# we can find maximum of a column using min function


print(min(iris$Sepal.Length))
print(min(iris$Sepal.Width))
print(min(iris$Petal.Length))
print(min(iris$Petal.Width))

```

*Finding the range of each column*
We use range() to finding the range of each column. It returns a vector containing the minimum and maximum of all the given arguments.

```{r}

print(range(iris$Sepal.Length))
print(range(iris$Sepal.Width))
print(range(iris$Petal.Length))
print(range(iris$Petal.Width))

```

*Finding variance of each column*
we use var() function to find the variance of each column
```{r}

print(var(iris$Sepal.Length))
print(var(iris$Sepal.Width))
print(var(iris$Petal.Length))
print(var(iris$Petal.Width))

```

*Finding standard deviation of each column*
We use sd() function to find the variance of each column
```{r}

print(sd(iris$Sepal.Length))
print(sd(iris$Sepal.Width))
print(sd(iris$Petal.Length))
print(sd(iris$Petal.Width))

```

**using summary function**
we can use summary() function to get all the above details in a single go.
```{r}
summary(iris)
```

# 2. R program to normalize the variables into 0 to 1 scale using min-max normalization

the formula to achieve min max normalization is 
y = (x-min)/(max-min)

$$y=\frac{x-min}{max-min}$$


```{r}
#dummy data
x = sample(-100:100,50)
print("original data")
print(x)
maximum = max(x)
minimum = min(x)
normalized = (x-minimum)/(maximum-minimum)
print("Normalized data")
print(normalized)
#using par function to fix multiple graphs in same plot

par(mfrow=c(1,2))
hist(x,breaks = 10, xlab = "Data",col = "lightblue", )
hist(normalized, breaks = 10, xlab = "Normalized data", col = "yellow")
```

#3.Generate histograms for any one variable and generate scatter plots for every pair of variables showing each species in different colour on iris dataset.

Let us use red, green, blue as the colours for 3 species 

